---
title: "Deep Learning Lab I (L05)"
author: "Data Science III (STAT 301-3)"
date: "May 12th, 2019"
output: html_document
---

# Overview

The goal of this lab is ensure that we are able to actually fit neural networks using the `keras` and `tensorflow` packages in R. We want to ensure that we can recreate the examples in provided in Chapters 2 and 3 in *Deep Learning with R*.

When creating the Rmd file for this assignment, you won't want all your code chunks to evaluate because it would take to long to run. You'll want to produce results, store them in a subdirectory, and then reference them in the Rmd file. You may need to use hidden chunks of code to read in results or use Rmd code to insert graphics that you want to display. **Make sure to thoroughly comment and describe what the code is doing in our own words.**

Author provides great [notebooks on github](https://github.com/jjallaire/deep-learning-with-r-notebooks) to go along with the book.

# Datasets 

All datasets are supplied by the `keras` package --- `mnist`, `imdb`, `rueters`, and `boston_housing`.

# Exercises

Please complete the following exercises. The document should be neatly formatted. 

#### Load Packages

```{r, message=FALSE}
# Loading package(s)
library(tensorflow)
use_python("/usr/bin/python")
# install_tensorflow(version = "1.12")
library(keras)
```
<br>

#### Exercise 1
Recreate the example provided in Section 2.1 (pg 25 - 29). See the [notebook](https://jjallaire.github.io/deep-learning-with-r-notebooks/notebooks/2.1-a-first-look-at-a-neural-network.nb.html) provided by the author.
```{r}
# load data
mnist <- dataset_mnist()
train_images <- mnist$train$x
train_labels <- mnist$train$y
test_images <- mnist$test$x
test_labels <- mnist$test$y
# have a look at the training data
str(train_images)
str(train_labels)
# have a look at the test data
str(test_images)
str(test_labels)
# build the network
network <- keras_model_sequential() %>% 
  layer_dense(units = 512, activation = "relu", input_shape = c(28 * 28)) %>% 
  layer_dense(units = 10, activation = "softmax")
# pick optimizer, loss function and monitor metrics
network %>% compile(
  optimizer = "rmsprop",
  loss = "categorical_crossentropy",
  metrics = c("accuracy")
)
# reshape the data
train_images <- array_reshape(train_images, c(60000, 28 * 28))
train_images <- train_images / 255
test_images <- array_reshape(test_images, c(10000, 28 * 28))
test_images <- test_images / 255
#categorically encode the labels
train_labels <- to_categorical(train_labels)
test_labels <- to_categorical(test_labels)
# train the network
network %>% fit(train_images, train_labels, epochs = 5, batch_size = 128)
# model performance on the test set
metrics <- network %>% evaluate(test_images, test_labels, verbose = 0)
metrics
```
Alternatively, students are welcome to work through the example provided on the [Keras in RStudio website](https://keras.rstudio.com/).
```{r}
# Preparing the Data
library(keras)
mnist <- dataset_mnist()
x_train <- mnist$train$x
y_train <- mnist$train$y
x_test <- mnist$test$x
y_test <- mnist$test$y
# reshape
x_train <- array_reshape(x_train, c(nrow(x_train), 784))
x_test <- array_reshape(x_test, c(nrow(x_test), 784))
# rescale
x_train <- x_train / 255
x_test <- x_test / 255
# one-hot encode the variables
y_train <- to_categorical(y_train, 10)
y_test <- to_categorical(y_test, 10)
# create a sequential model and then adding layers
model <- keras_model_sequential() 
model %>% 
  layer_dense(units = 256, activation = 'relu', input_shape = c(784)) %>% 
  layer_dropout(rate = 0.4) %>% 
  layer_dense(units = 128, activation = 'relu') %>%
  layer_dropout(rate = 0.3) %>%
  layer_dense(units = 10, activation = 'softmax')
# print the details of the model
summary(model)
# compile the model
model %>% compile(
  loss = 'categorical_crossentropy',
  optimizer = optimizer_rmsprop(),
  metrics = c('accuracy')
)
# train the model
history <- model %>% fit(
  x_train, y_train, 
  epochs = 30, batch_size = 128, 
  validation_split = 0.2
)
# plot object
plot(history)
# Evaluate the model’s performance on the test data
model %>% evaluate(x_test, y_test)
# Generate predictions on new data
model %>% predict_classes(x_test)
```
<br><br>

#### Exercise 2
Recreate the example provided in Section 3.4 (pg 59 - 69). See the [notebook](https://jjallaire.github.io/deep-learning-with-r-notebooks/notebooks/3.4-classifying-movie-reviews.nb.html) provided by the author.
```{r}
# load the dataset
library(keras)
imdb <- dataset_imdb(num_words = 10000)
c(c(train_data, train_labels), c(test_data, test_labels)) %<-% imdb
str(train_data[[1]])
train_labels[[1]]
max(sapply(train_data, max))
# decode the reviews back to English words
# word_index is a dictionary mapping words to an integer index
word_index <- dataset_imdb_word_index()
# We reverse it, mapping integer indices to words
reverse_word_index <- names(word_index)
names(reverse_word_index) <- word_index
# We decode the review; note that our indices were offset by 3
# because 0, 1 and 2 are reserved indices for "padding", "start of sequence", and "unknown".
decoded_review <- sapply(train_data[[1]], function(index) {
  word <- if (index >= 3) reverse_word_index[[as.character(index - 3)]]
  if (!is.null(word)) word else "?"
})
cat(decoded_review)
```

```{r}
# preparing the data
# turn the lists into tensors
vectorize_sequences <- function(sequences, dimension = 10000) {
  # Create an all-zero matrix of shape (len(sequences), dimension)
  results <- matrix(0, nrow = length(sequences), ncol = dimension)
  for (i in 1:length(sequences))
    # Sets specific indices of results[i] to 1s
    results[i, sequences[[i]]] <- 1
  results
}
# Our vectorized training data
x_train <- vectorize_sequences(train_data)
# Our vectorized test data
x_test <- vectorize_sequences(test_data)
str(x_train[1,])
# Our vectorized labels
y_train <- as.numeric(train_labels)
y_test <- as.numeric(test_labels)
```

```{r}
# buiding the network
model <- keras_model_sequential() %>% 
  layer_dense(units = 16, activation = "relu", input_shape = c(10000)) %>% 
  layer_dense(units = 16, activation = "relu") %>% 
  layer_dense(units = 1, activation = "sigmoid")
model %>% compile(
  optimizer = "rmsprop",
  loss = "binary_crossentropy",
  metrics = c("accuracy")
)
# custom compile
model %>% compile(
  optimizer = optimizer_rmsprop(lr = 0.001),
  loss = loss_binary_crossentropy,
  metrics = metric_binary_accuracy
) 
```

```{r}
# validating our approach
# set apart 10000 samples from the original training data
val_indices <- 1:10000
x_val <- x_train[val_indices,]
partial_x_train <- x_train[-val_indices,]
y_val <- y_train[val_indices]
partial_y_train <- y_train[-val_indices]
# train a new network from scratch for four epochs and then evaluate it on the test data
model <- keras_model_sequential() %>% 
  layer_dense(units = 16, activation = "relu", input_shape = c(10000)) %>% 
  layer_dense(units = 16, activation = "relu") %>% 
  layer_dense(units = 1, activation = "sigmoid")
model %>% compile(
  optimizer = "rmsprop",
  loss = "binary_crossentropy",
  metrics = c("accuracy")
)
model %>% fit(x_train, y_train, epochs = 4, batch_size = 512)
results <- model %>% evaluate(x_test, y_test)
results
# using a trained network to generate predictions on new data
model %>% predict(x_test[1:10,])
```
<br><br>

#### Exercise 3
Recreate the example provided in Section 3.5 (pg 70 - 76). See the [notebook](https://jjallaire.github.io/deep-learning-with-r-notebooks/notebooks/3.5-classifying-newswires.nb.html) provided by the author.
```{r}
# load dataset
library(keras)
reuters <- dataset_reuters(num_words = 10000)
c(c(train_data, train_labels), c(test_data, test_labels)) %<-% reuters
length(train_data)
length(test_data)
train_data[[1]]
# decode back to words
word_index <- dataset_reuters_word_index()
reverse_word_index <- names(word_index)
names(reverse_word_index) <- word_index
decoded_newswire <- sapply(train_data[[1]], function(index) {
  # Note that our indices were offset by 3 because 0, 1, and 2
  # are reserved indices for "padding", "start of sequence", and "unknown".
  word <- if (index >= 3) reverse_word_index[[as.character(index - 3)]]
  if (!is.null(word)) word else "?"
})
cat(decoded_newswire)
train_labels[[1]]
```

```{r}
# preparing the data
# vectorize the data
vectorize_sequences <- function(sequences, dimension = 10000) {
  results <- matrix(0, nrow = length(sequences), ncol = dimension)
  for (i in 1:length(sequences))
    results[i, sequences[[i]]] <- 1
  results
}
x_train <- vectorize_sequences(train_data)
x_test <- vectorize_sequences(test_data)
# one-hot encoding
to_one_hot <- function(labels, dimension = 46) {
  results <- matrix(0, nrow = length(labels), ncol = dimension)
  for (i in 1:length(labels))
    results[i, labels[[i]] + 1] <- 1
  results
}
one_hot_train_labels <- to_one_hot(train_labels)
one_hot_test_labels <- to_one_hot(test_labels)
# built-in way in Keras
one_hot_train_labels <- to_categorical(train_labels)
one_hot_test_labels <- to_categorical(test_labels)
```

```{r}
# building the network
model <- keras_model_sequential() %>% 
  layer_dense(units = 64, activation = "relu", input_shape = c(10000)) %>% 
  layer_dense(units = 64, activation = "relu") %>% 
  layer_dense(units = 46, activation = "softmax")
model %>% compile(
  optimizer = "rmsprop",
  loss = "categorical_crossentropy",
  metrics = c("accuracy")
)
# validate the approach
val_indices <- 1:1000
x_val <- x_train[val_indices,]
partial_x_train <- x_train[-val_indices,]
y_val <- one_hot_train_labels[val_indices,]
partial_y_train = one_hot_train_labels[-val_indices,]
# train the network for 20 epochs
history <- model %>% fit(
  partial_x_train,
  partial_y_train,
  epochs = 20,
  batch_size = 512,
  validation_data = list(x_val, y_val)
)
# display the loss and accuracy curves
plot(history)
```

```{r}
# train a new network from scratch for 9 epochs and then evaluate it on the test set
model <- keras_model_sequential() %>% 
  layer_dense(units = 64, activation = "relu", input_shape = c(10000)) %>% 
  layer_dense(units = 64, activation = "relu") %>% 
  layer_dense(units = 46, activation = "softmax")
  
model %>% compile(
  optimizer = "rmsprop",
  loss = "categorical_crossentropy",
  metrics = c("accuracy")
)
history <- model %>% fit(
  partial_x_train,
  partial_y_train,
  epochs = 9,
  batch_size = 512,
  validation_data = list(x_val, y_val)
)
results <- model %>% evaluate(x_test, one_hot_test_labels)
results
# compared with random baseline
test_labels_copy <- test_labels
test_labels_copy <- sample(test_labels_copy)
length(which(test_labels == test_labels_copy)) / length(test_labels)

# generating predictions on new data
predictions <- model %>% predict(x_test)
dim(predictions)
sum(predictions[1,])
which.max(predictions[1,])
```

```{r}
# what happens if intermediate layers less than 46-dimensional
model <- keras_model_sequential() %>% 
  layer_dense(units = 64, activation = "relu", input_shape = c(10000)) %>% 
  layer_dense(units = 4, activation = "relu") %>% 
  layer_dense(units = 46, activation = "softmax")
  
model %>% compile(
  optimizer = "rmsprop",
  loss = "categorical_crossentropy",
  metrics = c("accuracy")
)
history <- model %>% fit(
  partial_x_train,
  partial_y_train,
  epochs = 20,
  batch_size = 128,
  validation_data = list(x_val, y_val)
)
plot(history)
```
<br><br>

#### Exercise 4
Recreate the example provided in Section 3.6 (pg 76 - 83). See the [notebook](https://jjallaire.github.io/deep-learning-with-r-notebooks/notebooks/3.6-predicting-house-prices.nb.html) provided by the author.
```{r}
# load the data
library(keras)
dataset <- dataset_boston_housing()
c(c(train_data, train_targets), c(test_data, test_targets)) %<-% dataset
str(train_data)
str(test_data)
str(train_targets)
# preparing the data
mean <- apply(train_data, 2, mean)
std <- apply(train_data, 2, sd)
train_data <- scale(train_data, center = mean, scale = std)
test_data <- scale(test_data, center = mean, scale = std)
# build the network
build_model <- function() {
  model <- keras_model_sequential() %>% 
    layer_dense(units = 64, activation = "relu", 
                input_shape = dim(train_data)[[2]]) %>% 
    layer_dense(units = 64, activation = "relu") %>% 
    layer_dense(units = 1) 
    
  model %>% compile(
    optimizer = "rmsprop", 
    loss = "mse", 
    metrics = c("mae")
  )
}
```

```{r}
# k-fold cross validation 
k <- 4
indices <- sample(1:nrow(train_data))
folds <- cut(1:length(indices), breaks = k, labels = FALSE) 
num_epochs <- 100
all_scores <- c()
for (i in 1:k) {
  cat("processing fold #", i, "\n")
  # Prepare the validation data: data from partition # k
  val_indices <- which(folds == i, arr.ind = TRUE) 
  val_data <- train_data[val_indices,]
  val_targets <- train_targets[val_indices]
  
  # Prepare the training data: data from all other partitions
  partial_train_data <- train_data[-val_indices,]
  partial_train_targets <- train_targets[-val_indices]
  
  # Build the Keras model (already compiled)
  model <- build_model()
  
  # Train the model (in silent mode, verbose=0)
  model %>% fit(partial_train_data, partial_train_targets,
                epochs = num_epochs, batch_size = 1, verbose = 0)
                
  # Evaluate the model on the validation data
  results <- model %>% evaluate(val_data, val_targets, verbose = 0)
  all_scores <- c(all_scores, results$mean_absolute_error)
}  
all_scores
mean(all_scores)
```

```{r}
# try 500 epochs
# Some memory clean-up
k_clear_session()
num_epochs <- 500
all_mae_histories <- NULL
for (i in 1:k) {
  cat("processing fold #", i, "\n")
  
  # Prepare the validation data: data from partition # k
  val_indices <- which(folds == i, arr.ind = TRUE)
  val_data <- train_data[val_indices,]
  val_targets <- train_targets[val_indices]
  
  # Prepare the training data: data from all other partitions
  partial_train_data <- train_data[-val_indices,]
  partial_train_targets <- train_targets[-val_indices]
  
  # Build the Keras model (already compiled)
  model <- build_model()
  
  # Train the model (in silent mode, verbose=0)
  history <- model %>% fit(
    partial_train_data, partial_train_targets,
    validation_data = list(val_data, val_targets),
    epochs = num_epochs, batch_size = 1, verbose = 0
  )
  mae_history <- history$metrics$val_mean_absolute_error
  all_mae_histories <- rbind(all_mae_histories, mae_history)
}
# compute the average of the per-epoch MAE scores for all folds
average_mae_history <- data.frame(
  epoch = seq(1:ncol(all_mae_histories)),
  validation_mae = apply(all_mae_histories, 2, mean)
)
library(ggplot2)
ggplot(average_mae_history, aes(x = epoch, y = validation_mae)) + geom_line()
# get a clearer picture
ggplot(average_mae_history, aes(x = epoch, y = validation_mae)) + geom_smooth()
```

```{r}
# Build the final model (epochs = 70)
model <- build_model()
# Train the model (in silent mode, verbose=0)
model %>% fit(train_data, train_targets,
                epochs = 70, batch_size = 1, verbose = 0)
# Evaluate the model on the testing data
results <- model %>% evaluate(test_data, test_targets, verbose = 0)
results
```

